---
title: "QFISH"
author: "Payton Schlewitt"
date: "2024-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("tidyverse") #Only run once
#install.packages("usethis")  #Only run once
library("tidyverse")
library("usethis")
```

#This will vary from user to user depending on github credentials and should only be run when using a new device
use_git_config(user.name = "payschlew34", user.email = "pschlewitt@gmail.com")
#if on a new device, generate a new personal access token
gitcreds::gitcreds_set()


#Data sourced from the State of Queensland, Department of Agriculture, Fisheries

#Allow R to access our csv file
```{r}
commercial_line <- read.csv("../data/QFISH_Commercial_Line.csv")
#refining data set to exclude total values and 2024 because the data is still incomplete for that year
```

#Check our data!!
```{r}
head(commercial_line)
```
Year and Tonnes are characters
Need to remove comma from Tonnes calues 


```{r}
commercial_line$Tonnes
```


#Wranglin
```{r}
comm_line_2023 <-commercial_line |>
  slice(-35:-36) |>
  mutate(Tonnes =  str_replace(Tonnes, pattern = ",", replacement = ""))|>
  mutate_if(is.character, as.double)
head(comm_line_2023)
```
This removes the rows for total weight exported and the values for the year 2024 since the year does not have a complete dataset yet.
This also replaces the comma in values over 1,000 with no space so that Tonnes can be assigned as double instead of character.





#Creating a basic line plot with Tonnes vs Year
```{r}
ggplot(data = comm_line_2023, mapping = aes(x = CalendarYear, y = Tonnes)) +
  geom_point() +
  geom_line(group = 1) +
  scale_x_continuous(breaks = c(seq(1990, 2023, by = 5))) + 
  scale_y_continuous(breaks = seq(1500, 6500, by = 500)) +
  labs(title = "Yearly Commercial Line Fishery Catch", x = "Year", caption = "Data from the State of Queensland, Department of Agriculture, Fisheries")
```
```{r}
ggplot(
  data = comm_line_2023,mapping = aes(x=CalendarYear , y=Tonnes),)+
  geom_smooth()+
  
  #scale_x_discrete(breaks = seq(1990, 2023, by = 2)) + 
  scale_x_continuous(breaks = c(seq(1990, 2023, by = 5)))+
  scale_y_continuous(breaks = seq(1500, 6500, by = 500)) +
  labs(title = "Yearly Commercial Line Fishery Catch", x = "Year", caption = "Data from the State of Queensland, Department of Agriculture, Fisheries")
  
```


#Trying other graphs for remaining variables
```{r}
ggplot(data = comm_line_2023, mapping = aes(x = CalendarYear, y = Licences)) +
  geom_point() +
  geom_line(group = 1) +
  scale_x_continuous(breaks = c(seq(1990, 2023, by = 5))) +
  labs(x = "Year")
```

```{r}
#Days with catch reported
#Example: Two different fisherman go out and bring back fish on the same day. This is counted as 2 fishing days.
ggplot(data = comm_line_2023, mapping = aes(x = CalendarYear, y = Days)) +
  geom_point() +
  geom_line(group = 1) +
  scale_x_continuous(breaks = c(seq(1990, 2023, by = 5)))+
  labs(x = "Year")
```
