---
title: "MB5370_Mod4_Wrkshp3"
author: "Payton Schlewitt"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
library("tidyverse")
library("usethis")

#Example tables
```{r}
table1
```
#output:
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
```{r}
table2
```
#> # A tibble: 12 × 4
#>   country      year type           count
#>   <chr>       <int> <chr>          <int>
#> 1 Afghanistan  1999 cases            745
#> 2 Afghanistan  1999 population  19987071
#> 3 Afghanistan  2000 cases           2666
#> 4 Afghanistan  2000 population  20595360
#> 5 Brazil       1999 cases          37737
#> 6 Brazil       1999 population 172006362
#> # ... with 6 more rows
```{r}
table3
```
#> # A tibble: 6 × 3
#>   country      year rate             
#> * <chr>       <int> <chr>            
#> 1 Afghanistan  1999 745/19987071     
#> 2 Afghanistan  2000 2666/20595360    
#> 3 Brazil       1999 37737/172006362  
#> 4 Brazil       2000 80488/174504898  
#> 5 China        1999 212258/1272915272
#> 6 China        2000 213766/1280428583

#Example tables 2
# Compute rate per 10,000
```{r}
table1 %>% 
  mutate(rate = cases / population * 10000)
```
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
```{r}
table1 %>% 
  count(year, wt = cases)
```
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
```{r}
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))
```


#4.4.1 Exercises
#For each of the sample tables, describe what each observation and each column represents.
#Sketch out the processes you would use to calculate the rate for table2 and table3. You will need to perform four operations:
#1. Extract the number of TB cases per country per year
#2. Extract the matching population per country per year
#3. Divide cases by population, and multiply by 10,000
#4. Store back in the appropriate place

#I dont understand why this question is asking how to find rate for table 2 since table 2 has a column for rate.

#Table 1 has columns set up for country, year, cases and population
#Table 2 has columns set up for country, year, type (pop or cases) and count
#Table 3 has columns set up for country, year, and rate


#4.5 Pivoting data to make it tidy
#Using tidyr, we access two functions: pivot_longer() to lengthen data and pivot_wider() to widen data

#4.5.1 Lengthening datasets
```{r}
billboard
```
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...

#By sing pivot_longer(), we can tidy this table up by creating columns for week instead of rows
```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
``` 
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows

#cols specifies the columns we want to pivot, so for the previous table we wanted to pivot week, but this can be done with any column that is not a variable
#names_to names the variable stored in the column names
#values_to names the variable stored in the cells associated with week
#"week" and "rank" are in quotes because they are new variables nt defined when first running pivot_longer()
#performing pivot_longer() may create NA data values, but these can be eaasily removed by adding values_drop_na = TRUE
```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
```
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows
#NOTE: The total number of rows is now significantly reduced

#4.5.2 Pivoting longer
```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```
#We want the new dataset to have three variables:ID, measurement(column names), and value (the cell values)
#To do this, df needs to be pivoted longer
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```
#Output:
#> # A tibble: 6 × 3
#>   id    measurement value
#>   <chr> <chr>       <dbl>
#> 1 A     bp1           100
#> 2 A     bp2           120
#> 3 B     bp1           140
#> 4 B     bp2           115
#> 5 C     bp1           120
#> 6 C     bp2           125

#4.5.3 Widening datasets
#Using pivot_wider() we can condense observations scattered across multiple rows
#We are using a dataset from the Centers of Medicare and Medicaid
```{r}
cms_patient_experience
```
#Output:
#> # A tibble: 500 × 5
#>   org_pac_id org_nm                     measure_cd   measure_title   prf_rate
#>   <chr>      <chr>                      <chr>        <chr>              <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_1  CAHPS for MIPS...       63
#> 2 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_2  CAHPS for MIPS...       87
#> 3 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_3  CAHPS for MIPS...       86
#> 4 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_5  CAHPS for MIPS...       57
#> 5 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_8  CAHPS for MIPS...       85
#> 6 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_12 CAHPS for MIPS...       24
#> # ℹ 494 more rows

#Although the current columns would not make the best variable names, but for the sake of the exercise we are using 
#ulke pivot_longer(), pivot_wider() needs to be provided existing columns that define the vales and the column name
```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```
#Output:
#> # A tibble: 500 × 9
#>   org_pac_id org_nm                   measure_title   CAHPS_GRP_1 CAHPS_GRP_2
#>   <chr>      <chr>                    <chr>                 <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          63          NA
#> 2 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          87
#> 3 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 4 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 5 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> 6 0446157747 USC CARE MEDICAL GROUP ... CAHPS for MIPS...          NA          NA
#> # ℹ 494 more rows
#> # ℹ 4 more variables: CAHPS_GRP_3 <dbl>, CAHPS_GRP_5 <dbl>, ...

#To further refine this we need to specify which column or columns have values that uniquely identify each row, in this case they are the variables that start with "org"
```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```
#Output:
#> # A tibble: 95 × 8
#>   org_pac_id org_nm           CAHPS_GRP_1 CAHPS_GRP_2 CAHPS_GRP_3 CAHPS_GRP_5
#>   <chr>      <chr>                  <dbl>       <dbl>       <dbl>       <dbl>
#> 1 0446157747 USC CARE MEDICA...          63          87          86          57
#> 2 0446162697 ASSOCIATION OF ...          59          85          83          63
#> 3 0547164295 BEAVER MEDICAL ...          49          NA          75          44
#> 4 0749333730 CAPE PHYSICIANS...          67          84          85          65
#> 5 0840104360 ALLIANCE PHYSIC...          66          87          87          64
#> 6 0840109864 REX HOSPITAL INC          73          87          84          67
#> # ℹ 89 more rows
#> # ℹ 2 more variables: CAHPS_GRP_8 <dbl>, CAHPS_GRP_12 <dbl>
#Now everything is sorted by medical providers, much cleaner and easier to read

#4.5.4 Pivoting wider
```{r}
df2 <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```
#We want to take the names from the measurements column using names_from() and the values from the value column using values_from()
```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```
#Output:
#> # A tibble: 2 × 4
#>   id      bp1   bp2   bp3
#>   <chr> <dbl> <dbl> <dbl>
#> 1 A       100   120   105
#> 2 B       140   115    NA

#4.5.5 Exercises
#Why are pivot_longer() and pivot_wider() not perfectly symmetrical? Carefully consider the following example. 
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")

#These tables are not symmetrical because pivot_wider() is turning the years into rows and then assigning values, while pivot_longer() is turning the years into columns and then assigning values

#Why does this code fail?
```{r}
table4a %>% 
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
```
#> Error in `pivot_longer()`:
#> ! Can't subset columns past the end.
#> ℹ Locations 1999 and 2000 don't exist.
#> ℹ There are only 3 columns.

#ThIS code fails because the years are currently listed as (<dbl>) when they need to be listed as (<chr>)
#This can be fixed by changing the code to:
```{r}
table4a %>% 
  pivot_longer(c("1999", "2000"), names_to = "year", values_to = "cases")
```
#Consider the sample tibble below. Do you need to make it wider or longer? What are the variables?
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
#Make it wider. The variables are Gender and Pregnant, and the values are the count of each gender

#4.5.6 Separating and uniting data tables
#Using table 3 from earlier in the script, we want to separate the "rate" column into "cases" and "population" using searate()
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```
#Output:
#> # A tibble: 6 × 4
#>   country      year cases  population
#>   <chr>       <int> <chr>  <chr>     
#> 1 Afghanistan  1999 745    19987071  
#> 2 Afghanistan  2000 2666   20595360  
#> 3 Brazil       1999 37737  172006362 
#> 4 Brazil       2000 80488  174504898 
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
#Because the default for separate() turns our values into character (<chr>) types, we want those values to be numerical
#This is one by using convert = TRUE
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)
```
#Output:
#> # A tibble: 6 × 4
#>   country      year  cases population
#>   <chr>       <int>  <int>      <int>
#> 1 Afghanistan  1999    745   19987071
#> 2 Afghanistan  2000   2666   20595360
#> 3 Brazil       1999  37737  172006362
#> 4 Brazil       2000  80488  174504898
#> 5 China        1999 212258 1272915272
#> 6 China        2000 213766 1280428583
#unite() is the opposite tool to separate() and used to combine multiple columns into a single column
#For example, unite() is used here to combine the "century" and "year" columns
```{r}
table5 %>% 
  unite(new, century, year, sep = "") #sep = "" makes it so that there is no separation in the new combined column
#unite(new column name, column 1 to combine, column 2 to combine, sep = "" for none or "value/character")
```
#Output:
#> # A tibble: 6 × 3
#>   country     new   rate             
#>   <chr>       <chr> <chr>            
#> 1 Afghanistan 1999  745/19987071     
#> 2 Afghanistan 2000  2666/20595360    
#> 3 Brazil      1999  37737/172006362  
#> 4 Brazil      2000  80488/174504898  
#> 5 China       1999  212258/1272915272
#> 6 China       2000  213766/1280428583

#4.6 Handling missing values
```{r}
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4
)
```
#Missing columns can be filled using tidyr::fill()
```{r}
treatment |>
  fill(everything())
```
#> # A tibble: 4 × 3
#>   person           treatment response
#>   <chr>                <dbl>    <dbl>
#> 1 Derrick Whitmore         1        7
#> 2 Derrick Whitmore         2       10
#> 3 Derrick Whitmore         3       10
#> 4 Katherine Burke          1        4

#4.6.2 Fixed values
#Some values may represent a fixed or known value. This can be fixed using dplyr::coalesce() to replace NA values
```{r}
x <- c(1, 4, 5, 7, NA)
coalesce(x, 0)
```
#> [1] 1 4 5 7 0
#some values may be filled where they are actually an unknown value
#This can be fixed using dplyr::na_if()
#This is typically common in data generated by older software, represented by something like 99 or -999
```{r}
x <- c(1, 4, 5, 7, -99)
na_if(x, -99)
```
#> [1]  1  4  5  7 NA
#A special type of missing value is NaN (Not a NUmber) and can be distinguished from NA by using is.nan(x)
```{r}
x <- c(NA, NaN)
x * 10
#> [1]  NA NaN
x == 1
#> [1] NA NA
is.na(x)
#> [1] TRUE TRUE
```
#4.6.4 Implicit missing values
#To demonstrate what a missing implicit value is, we'll use this example below:
```{r}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```
#This dataset is missing the price for the 4th quarter in 2020, as well as the entire 1st quarter of 2021
#By turning implicit values in explicit, we give ourselves something physical to work with.
#By widening the data using pivot_wider() we can turn the implicit values from before into explicit values
```{r}
stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  )
```
#Output:
#> # A tibble: 2 × 5
#>    year   `1`   `2`   `3`   `4`
#>   <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1  2020  1.88  0.59  0.35 NA   
#> 2  2021 NA     0.92  0.17  2.66

#4.7.1 CSV files
#Example CSV
Student ID,Full Name,favourite.food,mealPlan,AGE 1,Sunil Huffmann,Strawberry yoghurt,Lunch only,4 2,Barclay Lynn,French fries,Lunch only,5 3,Jayendra Lyne,N/A,Breakfast and lunch,7 4,Leon Rossini,Anchovies,Lunch only, 5,Chidiegwu Dunkel,Pizza,Breakfast and lunch,five 6,Güvenç Attila,Ice cream,Lunch only,6
#The first line is the header row, which gives the columns names
#The following lines provide the data
#columns are separated by commas so each data value also has a comma separating it
#To have R read a CSV file, use read_csv(path for file)
#For a file with the name students.csv:
? read_csv
students <- read_csv("C://data/students.csv")
#> Rows: 6 Columns: 5
students <- read_csv("https://pos.it/r4ds-students-csv")

